---
title: "Data Manipulation"
author: "Devon Morgan"
date: "9/20/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Import csv files

Import csvs. 

```{r}
litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)
```

## Select variables

... by listing using the `select` function: 

```{r}
select(litters_data, group, litter_number, gd0_weight)

select(litters_data, litter_number, gd0_weight, pups_survive)

```

... by specifying a range: 

```{r}
# using range alone 
select(litters_data, gd_of_birth:pups_survive)

# can combine the function for range and listing together:
select(litters_data, group, gd_of_birth:pups_survive)
```

... by saying what you want to remove: 

```{r}
select(litters_data, -litter_number)
```

... can combine the selection with renaming variables. Do this by putting the new name first: 

```{r}
select(litters_data, group, litter_number, gest_day_0_weight = gd0_weight)

# can also do this by using the `rename` function: 

rename(litters_data, gest_day_0_weight = gd0_weight)
```

The `?select_helpers` functions are helpful for selecting variables based on certain criteria. Using select helpers: 

```{r}
select(litters_data, starts_with("gd"))
select(litters_data, litter_number, starts_with("gd"))

# using the everything function -- helps reorganize columns without selecting: 
select(litters_data, litter_number, everything())

```

### Variable selecting in Pups dataset

Learning assessment: 

```{r}
select(pups_data, litter_number, sex, pd_ears)
```

## Filter observations

Eliminates some rows from the dataset. Using logical operators to select these. `==` is interpreted to mean exactly equal, and does not function as assigning a new variable. Using the `filter` function:

```{r}
filter(litters_data, gd0_weight < 25)
filter(litters_data, gd0_weight >= 25)

filter(litters_data, pups_born_alive == 8)

# is it true that gd0 weight is missing?
filter(litters_data, is.na(gd0_weight))

# excluding when weight is NA: 

filter(litters_data, !is.na(gd0_weight))

# keep only some groups: 

filter(litters_data, group =="Low8" | group == "Low7")
filter(litters_data, group %in% c("Low7", "Low8"))

filter(litters_data, pups_born_alive %in% 4:6)

# listing both functions as an "AND" operator
filter(litters_data, pups_born_alive %in% 4:7, !is.na(gd0_weight))
```


Learning assessment: 

```{r}
filter(pups_data, sex == 1)

filter(pups_data, sex == 2, pd_walk < 11)
```

## Mutate function

Create variables. Inside mutate, can create multiple variables at the same time. 

```{r}
mutate(litters_data,
    wt_gain = gd18_weight - gd0_weight,
    group = tolower(group)
       )

mutate(litters_data,
      wt_gain = gd18_weight - gd0_weight,
      wt_gain_squared = wt_gain^2)

```

## Arrange function

Just reordering data, but not changing anything. 

Arrange the data...

```{r}
arrange(litters_data, gd0_weight)

arrange(litters_data, pups_born_alive, gd0_weight)

arrange(litters_data, desc(pups_born_alive), gd0_weight)
```

## Piping!!

Look at intermediate object approach: 


```{r}
litters_data_raw = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars = mutate(litters_data_selected_cols, 
  wt_gain = gd18_weight - gd0_weight,
  group = tolower(group))
litters_data_with_vars
```

Use nested function calls. This code does the same as the previous code chunk:

```{r}
litters_data_clean = 
  mutate(
    select(
      janitor::clean_names(
        read_csv("./data/FAS_litters.csv", col_types = "ccddiiii")
        ), 
    -pups_survive
    ),
  wt_gain = gd18_weight - gd0_weight,
  group = tolower(group)
  )
litters_data_clean
```

Using piping. Executes each step sequentially: 

```{r}
read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>% 
  select(-pups_survive) %>% 
  mutate(
    wt_gain = gd18_weight - gd0_weight, 
    group = tolower(group)
  ) %>% 
  filter(., !is.na(gd0_weight))


```


Illustrate placeholder use with `lm`

```{r}
read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>% 
  select(-pups_survive) %>% 
  mutate(
    wt_gain = gd18_weight - gd0_weight, 
    group = tolower(group)
  ) %>% 
  filter(., !is.na(gd0_weight)) %>% 
  lm(gd18_weight ~ gd0_weight, data = .)

```

```{r}
litters_data =
read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>% 
  select(-pups_survive) %>% 
  mutate(
    wt_gain = gd18_weight - gd0_weight, 
    group = tolower(group)
    ) %>% 
  filter(., !is.na(gd0_weight)) 

```